
output:
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5,
  fig.align = "center"
)
```

# Setup & Package Installation

Install packages if they are not already available, then load them.

```{r packages}
# Uncomment the lines below only the FIRST time you run this notebook
# install.packages(c("quantmod", "TTR", "ggplot2", "plotly",
#                    "forecast", "PerformanceAnalytics", "dplyr",
#                    "zoo", "xts", "lubridate", "gridExtra"))

library(quantmod)            # Fetching & handling financial time-series
library(TTR)                 # Technical trading rules (SMA, EMA, runSD …)
library(ggplot2)             # Grammar-of-graphics visualizations
library(plotly)              # Interactive charts
library(forecast)            # ARIMA modelling & forecasting
library(PerformanceAnalytics) # Portfolio analytics
library(dplyr)               # Data wrangling
library(zoo)                 # Irregular time-series helpers
library(xts)                 # Extensible time-series objects
library(lubridate)           # Date arithmetic
library(gridExtra)           # Multi-panel ggplot layouts
```

---

# Fetching & Preprocessing Data

## Single Stock — Apple (AAPL)

`quantmod::getSymbols()` downloads OHLCV data from Yahoo Finance and stores it as an **xts** object in the global environment.

```{r fetch-aapl}
# Download AAPL data for 2020–2021
getSymbols("AAPL", from = "2020-01-01", to = "2022-01-01", auto.assign = TRUE)

# Adjust closing prices for stock splits and dividend payouts
adjusted_data <- adjustOHLC(AAPL, use.Adjusted = TRUE)

# Inspect the first few rows
head(adjusted_data)
```

**Key columns produced by `getSymbols`:**

| Column | Meaning |
|--------|---------|
| `AAPL.Open` | Opening price for the day |
| `AAPL.High` | Highest intraday price |
| `AAPL.Low` | Lowest intraday price |
| `AAPL.Close` | Closing price |
| `AAPL.Volume` | Total shares traded |
| `AAPL.Adjusted` | Close price adjusted for corporate actions |

---

## Handling Missing Values

Real-world financial data can have gaps (holidays, halted trading).

```{r missing-values}
# Count missing values per column
na_count <- colSums(is.na(adjusted_data))
cat("Missing values per column:\n")
print(na_count)

# Forward-fill any gaps (last observed value carried forward)
adjusted_data_clean <- na.locf(adjusted_data)
cat("\nMissing values after forward-fill:\n")
print(colSums(is.na(adjusted_data_clean)))
```

---

# Exploratory Data Analysis

## Distribution of Daily Returns

The **daily return** measures the percentage change from one trading day to the next.  
$$r_t = \frac{P_t - P_{t-1}}{P_{t-1}}$$

```{r daily-returns}
# Compute daily returns from adjusted close
returns <- dailyReturn(Cl(adjusted_data_clean))

# Quick summary statistics
summary(returns)
cat("\nStandard deviation of daily returns:", round(sd(returns, na.rm = TRUE), 5), "\n")
```

```{r histogram-returns, fig.cap="Distribution of AAPL daily returns (2020-2021)"}
# Convert xts to data frame for ggplot
returns_df <- data.frame(
  Date         = index(returns),
  daily.returns = as.numeric(returns)
)

ggplot(returns_df, aes(x = daily.returns)) +
  geom_histogram(bins = 60, fill = "#1a6faf", color = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, color = "red", linewidth = 0.8, linetype = "dashed") +
  geom_vline(xintercept = mean(returns_df$daily.returns, na.rm = TRUE),
             color = "orange", linewidth = 0.8) +
  labs(
    title    = "Histogram of AAPL Daily Returns (2020–2021)",
    subtitle = "Red dashed = zero  |  Orange = mean return",
    x        = "Daily Return",
    y        = "Frequency"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

# Moving Average Analysis

## Fetching Multiple Stocks

```{r fetch-multi}
# Download AAPL, GOOG, and MSFT simultaneously
getSymbols(c("AAPL", "GOOG", "MSFT"),
           from = "2020-01-01", to = "2022-01-01",
           auto.assign = TRUE)

# Adjust for splits & dividends
aapl_adj <- adjustOHLC(AAPL, use.Adjusted = TRUE)
goog_adj <- adjustOHLC(GOOG, use.Adjusted = TRUE)
msft_adj <- adjustOHLC(MSFT, use.Adjusted = TRUE)

cat("AAPL rows:", nrow(aapl_adj),
    "  GOOG rows:", nrow(goog_adj),
    "  MSFT rows:", nrow(msft_adj), "\n")
```

## Computing Simple Moving Averages (SMA)

The **Simple Moving Average** smooths noisy price data by averaging over a rolling window of *n* days:
$$\text{SMA}(n)_t = \frac{1}{n}\sum_{i=0}^{n-1} P_{t-i}$$

```{r sma}
# 50-day SMA for each stock
aapl_adj$SMA50 <- SMA(Cl(aapl_adj), n = 50)
goog_adj$SMA50 <- SMA(Cl(goog_adj), n = 50)
msft_adj$SMA50 <- SMA(Cl(msft_adj), n = 50)

# Also compute a 20-day SMA for a faster signal
aapl_adj$SMA20 <- SMA(Cl(aapl_adj), n = 20)
```

## Plotting AAPL with Moving Averages

```{r plot-aapl-sma, fig.cap="AAPL adjusted close with 20-day and 50-day SMA"}
aapl_df <- data.frame(
  Date  = index(aapl_adj),
  Close = as.numeric(Cl(aapl_adj)),
  SMA20 = as.numeric(aapl_adj$SMA20),
  SMA50 = as.numeric(aapl_adj$SMA50)
)

ggplot(aapl_df, aes(x = Date)) +
  geom_line(aes(y = Close, color = "Close Price"), linewidth = 0.6, alpha = 0.7) +
  geom_line(aes(y = SMA20, color = "20-day SMA"), linewidth = 0.9) +
  geom_line(aes(y = SMA50, color = "50-day SMA"), linewidth = 0.9) +
  scale_color_manual(values = c("Close Price" = "#aec6e8",
                                 "20-day SMA"  = "#ff7f0e",
                                 "50-day SMA"  = "#d62728")) +
  labs(title    = "AAPL Adjusted Closing Prices with Moving Averages",
       subtitle = "2020–2021 | Golden/Death Cross signals visible",
       x = "Date", y = "Adjusted Close (USD)", color = "Legend") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"))
```

## Plotting GOOG with 50-day SMA

```{r plot-goog-sma, fig.cap="GOOG adjusted close with 50-day SMA"}
goog_df <- data.frame(
  Date  = index(goog_adj),
  Close = as.numeric(Cl(goog_adj)),
  SMA50 = as.numeric(goog_adj$SMA50)
)

ggplot(goog_df, aes(x = Date)) +
  geom_line(aes(y = Close, color = "Close Price"), linewidth = 0.6, alpha = 0.7) +
  geom_line(aes(y = SMA50, color = "50-day SMA"), linewidth = 0.9) +
  scale_color_manual(values = c("Close Price" = "#98df8a", "50-day SMA" = "#2ca02c")) +
  labs(title = "GOOG Adjusted Closing Prices with 50-day SMA",
       x = "Date", y = "Adjusted Close (USD)", color = "Legend") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

## Plotting MSFT with 50-day SMA

```{r plot-msft-sma, fig.cap="MSFT adjusted close with 50-day SMA"}
msft_df <- data.frame(
  Date  = index(msft_adj),
  Close = as.numeric(Cl(msft_adj)),
  SMA50 = as.numeric(msft_adj$SMA50)
)

ggplot(msft_df, aes(x = Date)) +
  geom_line(aes(y = Close, color = "Close Price"), linewidth = 0.6, alpha = 0.7) +
  geom_line(aes(y = SMA50, color = "50-day SMA"), linewidth = 0.9) +
  scale_color_manual(values = c("Close Price" = "#ffbb78", "50-day SMA" = "#ff7f0e")) +
  labs(title = "MSFT Adjusted Closing Prices with 50-day SMA",
       x = "Date", y = "Adjusted Close (USD)", color = "Legend") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

---

# Volatility Analysis

Volatility measures **how much prices fluctuate** over time. High volatility = higher risk and potential reward.

We use a **rolling standard deviation** over a 20-day window:
$$\sigma_t^{(20)} = \text{std}(P_{t-19}, \ldots, P_t)$$

```{r rolling-vol}
# 20-day rolling standard deviation for each stock
aapl_adj$Volatility <- runSD(Cl(aapl_adj), n = 20)
goog_adj$Volatility <- runSD(Cl(goog_adj), n = 20)
msft_adj$Volatility <- runSD(Cl(msft_adj), n = 20)
```

## Individual Volatility Plots

```{r plot-aapl-vol, fig.cap="AAPL 20-day rolling volatility"}
aapl_vol_df <- data.frame(
  Date       = index(aapl_adj),
  Volatility = as.numeric(aapl_adj$Volatility)
)

ggplot(aapl_vol_df, aes(x = Date, y = Volatility)) +
  geom_line(color = "#1a6faf", linewidth = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "red", linewidth = 0.6, alpha = 0.2) +
  labs(title    = "AAPL — 20-Day Rolling Volatility",
       subtitle = "Shaded band = LOESS smoothing confidence interval",
       x = "Date", y = "Rolling Std. Dev (USD)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## Comparative Volatility — All Three Stocks

```{r compare-volatility, fig.cap="Comparative volatility: AAPL vs GOOG vs MSFT"}
# Merge into one data frame
vol_df <- data.frame(
  Date = index(aapl_adj),
  AAPL = as.numeric(aapl_adj$Volatility),
  GOOG = as.numeric(goog_adj$Volatility),
  MSFT = as.numeric(msft_adj$Volatility)
)

# Reshape to long format
vol_long <- tidyr::pivot_longer(vol_df, cols = c("AAPL", "GOOG", "MSFT"),
                                names_to = "Stock", values_to = "Volatility")

ggplot(vol_long, aes(x = Date, y = Volatility, color = Stock)) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  scale_color_manual(values = c("AAPL" = "#1a6faf",
                                 "GOOG" = "#2ca02c",
                                 "MSFT" = "#ff7f0e")) +
  labs(title    = "Comparative 20-Day Rolling Volatility",
       subtitle = "AAPL vs GOOG vs MSFT (2020–2021)",
       x = "Date", y = "Rolling Std. Dev (USD)", color = "Stock") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

---

# Correlation Analysis

```{r correlation}
# Align closing prices into a single xts matrix
prices <- merge(Cl(aapl_adj), Cl(goog_adj), Cl(msft_adj))
colnames(prices) <- c("AAPL", "GOOG", "MSFT")

# Compute pairwise correlation of daily returns
ret_matrix <- diff(log(prices))  # log returns
cor_matrix  <- cor(ret_matrix, use = "complete.obs")
cat("Pairwise Correlation Matrix (log returns):\n")
round(cor_matrix, 4)
```

```{r cor-heatmap, fig.cap="Correlation heatmap of log returns"}
cor_df <- reshape2::melt(cor_matrix)
ggplot(cor_df, aes(x = Var1, y = Var2, fill = value, label = round(value, 3))) +
  geom_tile(color = "white") +
  geom_text(size = 5, fontface = "bold") +
  scale_fill_gradient2(low = "#d73027", mid = "#ffffbf", high = "#1a9850",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Heatmap — AAPL, GOOG, MSFT Daily Log Returns",
       x = NULL, y = NULL, fill = "ρ") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        axis.text  = element_text(size = 12))
```

---

# Predictive Modelling — ARIMA Forecasting

## What is ARIMA?

**ARIMA(p, d, q)** — AutoRegressive Integrated Moving Average — is one of the most widely used time-series forecasting models:

| Parameter | Role |
|-----------|------|
| **p** | Number of autoregressive lags |
| **d** | Degree of differencing (to achieve stationarity) |
| **q** | Number of moving-average lags |

`forecast::auto.arima()` automatically selects optimal (p, d, q) using AIC/BIC.

## Fitting the Model on AAPL

```{r arima-fit}
# Extract the closing price as a plain numeric vector
aapl_close <- as.numeric(Cl(aapl_adj))

# auto.arima selects optimal order by minimising AIC
arima_model <- auto.arima(aapl_close, seasonal = FALSE, stepwise = FALSE,
                           approximation = FALSE, trace = FALSE)
summary(arima_model)
```

## Generating a 30-Day Forecast

```{r arima-forecast, fig.cap="AAPL 30-day forecast using ARIMA"}
# Forecast 30 trading days ahead
fc <- forecast(arima_model, h = 30, level = c(80, 95))

# Build a tidy data frame for plotting
n       <- length(aapl_close)
dates   <- index(aapl_adj)
fc_dates <- seq(tail(dates, 1) + 1, by = "day", length.out = 30)

hist_df <- data.frame(Date = dates, Price = aapl_close, Type = "Historical")
fc_df   <- data.frame(
  Date   = fc_dates,
  Price  = as.numeric(fc$mean),
  Lo80   = as.numeric(fc$lower[, 1]),
  Hi80   = as.numeric(fc$upper[, 1]),
  Lo95   = as.numeric(fc$lower[, 2]),
  Hi95   = as.numeric(fc$upper[, 2])
)

ggplot() +
  geom_line(data = tail(hist_df, 120),
            aes(x = Date, y = Price), color = "#1a6faf", linewidth = 0.7) +
  geom_ribbon(data = fc_df, aes(x = Date, ymin = Lo95, ymax = Hi95),
              fill = "#d62728", alpha = 0.15) +
  geom_ribbon(data = fc_df, aes(x = Date, ymin = Lo80, ymax = Hi80),
              fill = "#d62728", alpha = 0.30) +
  geom_line(data = fc_df, aes(x = Date, y = Price),
            color = "#d62728", linewidth = 0.9, linetype = "dashed") +
  labs(title    = "AAPL — ARIMA 30-Day Price Forecast",
       subtitle = "Dark band = 80% CI  |  Light band = 95% CI",
       x = "Date", y = "Adjusted Close (USD)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

# Interactive Visualization with Plotly

```{r plotly-chart, fig.cap="Interactive AAPL price chart (hover for details)"}
aapl_plot_df <- data.frame(
  Date  = index(aapl_adj),
  Open  = as.numeric(Op(aapl_adj)),
  High  = as.numeric(Hi(aapl_adj)),
  Low   = as.numeric(Lo(aapl_adj)),
  Close = as.numeric(Cl(aapl_adj))
)

plot_ly(aapl_plot_df,
        x    = ~Date,
        type = "candlestick",
        open = ~Open, close = ~Close,
        high = ~High, low   = ~Low,
        name = "AAPL") %>%
  layout(
    title   = "AAPL — Candlestick Chart (2020–2021)",
    xaxis   = list(title = "Date", rangeslider = list(visible = TRUE)),
    yaxis   = list(title = "Price (USD)")
  )
```

---

# Summary & Key Takeaways

```{r summary-stats}
# Compute key metrics for each stock
stocks_list <- list(AAPL = aapl_adj, GOOG = goog_adj, MSFT = msft_adj)

summary_table <- lapply(names(stocks_list), function(ticker) {
  s      <- stocks_list[[ticker]]
  ret    <- dailyReturn(Cl(s))
  data.frame(
    Stock          = ticker,
    Start_Price    = round(as.numeric(first(Cl(s))), 2),
    End_Price      = round(as.numeric(last(Cl(s))), 2),
    Total_Return_pct = round((as.numeric(last(Cl(s))) / as.numeric(first(Cl(s))) - 1) * 100, 2),
    Ann_Volatility_pct = round(sd(ret, na.rm = TRUE) * sqrt(252) * 100, 2),
    Max_Daily_Return  = round(max(ret, na.rm = TRUE) * 100, 3),
    Min_Daily_Return  = round(min(ret, na.rm = TRUE) * 100, 3)
  )
})

final_summary <- do.call(rbind, summary_table)
knitr::kable(final_summary,
             caption = "Summary Statistics — AAPL, GOOG, MSFT (2020–2021)",
             align = "c")
```
